EDIpack2: interoperable Lanczos-based solver for generic quantum impurity problems

### 2025-04-03 edipack2.tex
Update draft. Completed sec.2, the main part

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or
%% 5p,twocolumn, preprint, review
\documentclass[final,3p,10pt]{elsarticle}

@@

@@

\subsection{Reduced impurity density matrix}\label{sSecRDM}
This version of the \NAME library features the calculation of the
impurity Reduced Density Matrix (iRDM), enabling the analysis of
entanglement properties of the quantum impurity for any value of {\tt
  ed\_mode}.
To simplify the discussion in this section we take the zero
temperature limit and consider the presence of a non-degenerate
groundstate $\ket{\psi}$ populating the {\tt state\_list}.
The generalization to the presence of degenerate states or the finite
temperature regime are straightforward.
The groundstate belongs to a given symmetry sector $\SS$ and it
represented in the Fock basis as:  $\ket{\psi} = \sum_I a_I \ket{I}$. 
The density matrix $\rho$ is defined as:
$$
\rho = \ket{\psi}\bra{\psi} = \sum_{nm=1}^{4^{N_s}} a^*_ma_n \ket{n}\bra{m} =\sum_{nm=1}^{4^{N_s}} \rho_{nm} \ket{n}\bra{m} 
$$
The iRDM is obtained from this by tracing out the bath degrees of
freedom: $\rho_{imp}=Tr_{bath}\rho$. The summation appearing in the
previous expressions becomes quickly unmaneagable with the system
size. To overcome this issue we proceed in three steps. 
We use quantum numbers conservation which
reduces extrema of the summation to the size $D_S$ of the sector $S$
to which the groundstate belongs;
Next we split the bits representation of any state $\ket{n}$ into
their spin-dependent impurity and bath parts as $\ket{n} = \ket{
  i_\uparrow\,  b_\uparrow\, i_\downarrow \, b_\downarrow}$. Each bit
set span a space of dimension $D^\a_{imp}$ or $D^\s_{bath}$. 
Finally, we store into the {\tt sparse\_map} ${\cal S}$, for every impurity configuration 
$i_\s =1,\dots,D^\s_{imp}$ (key) the corresponding bath states $j\in[1,\dots,D^\s_{bath}]$
(values) which contribute to that key, as detailed in
\secu{CodeSparseMap}.
We indicate with ${\cal D}^\s_{i}$ the number of values for the key $i_\s$.
The fast algorithm for the evaluation of the iRDM are outlined below
for the two cases {\bf normal} and {\bf superc/nonsu2}.

\paragraph{{\bf normal}}

@@

\subsection{Bath Functions}\label{sSecFunc}

\subsection{Bath Optimization}\label{sSecFit}
In the DMFT context the bath is updated using a suitable 
optimization of the bath parameters against a given realization of the Weiss field
${\GG^{-1}_{0}}_{\a\b}(z)$, or the corresponding Delta function
$\Delta_{\a\b}(z)= (z + \mu)\delta_{\a\b} - H^{loc}_{\a\b}-
{\GG^{-1}_0}_{\a\b}(z)$, $z\in\CCC$, determined by the self-consistency procedure~\cite{Georges1996RMP}.
This step corresponds to a projection of the Weiss
field (or the Delta function) onto the space of the Anderson quantum
impurity model functions
$\GG^{And}_0(z) =
[(z + \mu)\delta_{\a\b} - H^{loc}_{\a\b} -
\Delta_{\a\b}(z)]^{-1}$, with $\Delta^{And}_{\a\b}(z)=\sum_\nu
{V^\nu_{\a}}(z-h^\nu_{\a\b})^{-1}V^\nu_{\b}$, spanned by the
variational paramaters $\{V,h\}$.

This  optimization can be performed using different
  algorithms~\cite{Garcia2004PRL,Taranto2012PRB}. 
  In \NAME\ we offer to perform this task using a  
conjugate gradient (CG) minimization of the cost function:
$$
\chi = \sum_{n=1}^{L_{fit}}\frac{1}{W_n}\sum_{\a\b}|X_{\a\b}(i\omega_n) - X_{\a\b}^{And}(i\omega_n;\{V,h\})|^q
$$
where $X_{\a\b}={\GG_{0}}_{\a\b},\,\Delta_{\a\b}$ are the user
supplied local functions  and
$X_{\a\b}^{And}={\GG_0}^{And}_{\a\b},\,\Delta^{And}_{\a\b}$ the
Anderson impurity model function, defined above.
The fit procedures are contains in the module \texttt{BATH\_FIT}.  
In order to exploit regularity of the functions the fit is 
performed using imaginary Matsubara frequency. 
The actual form of $X_{\a\b}$ is controlled by the input parameter
\texttt{cg\_Scheme=Weiss,Delta}.

Given the importance of the fit for the convergence of DMFT calculations, in \NAME\ we
introduced different specific control parameters. This aims to give
the user the possibility of fine tuning the fit and adapt its behavior to
specific situations.
Two distinct CG algorithms can be used in the library, as controlled by the input variable
\texttt{cg\_method=0,1}. The value \texttt{0} corresponds to a
Fletcher-Reeves-Polak-Ribiere minimisation algorithm, adapted from
Numerical Recipes~\cite{NumRec77}. In order to guarantee
back-compatibility with respect to the literature, we also provide the
possibility of using the minimization procedure published in
Ref.\onlinecite{Georges1996RMP} which has been largely used in the DMFT
community.

@@

\subsection{Input/Output}\label{sSecIO}

>>>>

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or
%% 5p,twocolumn, preprint, review
\documentclass[preprint,3p,10pt]{elsarticle}

@@

\usepackage[inline]{showlabels}

@@

\subsection{Reduced impurity density matrix}\label{sSecRDM}
This version of the \NAME library features the calculation of the
impurity Reduced Density Matrix (iRDM), enabling the analysis of
entanglement properties of the quantum impurity for any value of {\tt
  ed\_mode}.
To simplify the discussion in this section we take the zero
temperature limit and consider the presence of a non-degenerate
groundstate $\ket{\psi}$ populating the {\tt state\_list}.
The generalization to the presence of degenerate states or the finite
temperature regime are straightforward.
The groundstate belongs to a given symmetry sector $\SS$ and it
represented in the Fock basis as:  $\ket{\psi} = \sum_I a_I \ket{I}$. 
The density matrix $\rho$ is defined as:
$$
\rho = \ket{\psi}\bra{\psi} = \sum_{nm=1}^{4^{N_s}} a^*_ma_n \ket{n}\bra{m} =\sum_{nm=1}^{4^{N_s}} \rho_{nm} \ket{n}\bra{m} 
$$
The iRDM is obtained from this by tracing out the bath degrees of
freedom: $\rho^{i}=Tr_{b}\rho$. The summation appearing in the
previous expressions becomes quickly unmaneagable with the system
size. To overcome this issue we implemented a fast algorithm
leveraging on symmetry sector and map sparsity.
Using quantum number conservation the summation extrema get reducedto
the size $D_S$ of the sector $S$ to which the groundstate (or in
general any eigenstate) belongs. 
Next we analyze the groundstate expansion on the Fock states, splitting the bit representation of any state $\ket{I}$ into
their spin-dependent impurity and bath parts as $\ket{I} = \ket{
  i_\uparrow\,  b_\uparrow\, i_\downarrow \, b_\downarrow}$. Each bit
set span a space of dimension $D^\a_{imp}$ or $D^\s_{bath}$. 
Finally, we store into the {\tt sparse\_map} ${\cal S}$, for every impurity configuration 
$i_\s =1,\dots,D^\s_{imp}$ (key) the corresponding bath states $j\in[1,\dots,D^\s_{bath}]$
(values) which contribute to that key, as detailed in
\secu{CodeSparseMap}.
We indicate with ${\cal D}^\s_{i}$ the number of values for the key $i_\s$.
The fast summation algorithm for the evaluation of the iRDM takes on a
different form for the two cases {\bf normal} and {\bf superc/nonsu2},
as outlined below.

\paragraph{{\bf normal}}
In this case we take advantage of the Fock and sector spin resolved
factorization discussed in \secu{sSecQNs}. As such we can also split
impurity and bath bits independently for each spin configuration:
$$
\ket{\vec{n}} =
\ket{\vec{n}_\up}\otimes \ket{\vec{n}_\dw} =
\ket{\vec{i}_\up}\ket{\vec{b}_\up}\otimes \ket{\vec{i}_\dw}\ket{\vec{b}_\dw}
$$
In addition it should be noted that creation (destruction) operators at
position $p$ for spin $\sigma$  $c^+_{p\sigma}$ ($c^+_{p\sigma}$) do
not operate on the opposite spin so that they pass through without
taking any fermionic sign. Finally, we observe that local iRDM admits
only diagonal terms in the {\bf normal} case. Thus we obtain for the iRDM:
\begin{equation}
  \label{iRDMnormal}
  \begin{aligned}
  \rho^i &=
  \sum_{\sigma=\up\dw}\sum_{b_\sigma=1}^{D_{b_\sigma}}
  \bra{b_\up}\otimes\bra{b_\dw}
    \rho
    \ket{b_\dw}\otimes\ket{b_\up}    \cr
    %
    &=\sum_{\sigma=\up\dw}
    \sum_{b_\sigma=1}^{D_{b_\sigma}}
    \sum_{i_\sigma=1}^{D_{i_\sigma}}
    \sum_{p_\sigma=1}^{D_{p_\sigma}}
    \sum_{j_\sigma=1}^{D_{j_\sigma}} 
    \sum_{q_\sigma=1}^{D_{q_\sigma}}
    C_{i,p}\,C_{j,q}\,a_{i_\up p_\up i_\dw p_\dw} a^*_{j_\up q_\up j_\dw q_\dw}
    \langle b_\up| p_\up\rangle| i_\up\rangle\langle j_\up |\langle q_\up| b_\up\rangle   \otimes \langle b_\dw| p_\dw \rangle|i_\dw\rangle \langle j_\dw  |\langle q_\dw| b_\dw\rangle
    \cr
&=
    \sum_{b_\sigma=1}^{D_{b_\sigma}}
    \sum_{i_\sigma=1}^{D_{i_\sigma}}
    a_{i_\up b_\up i_\dw b_\dw} a^*_{i_\up b_\up i_\dw b_\dw}
    \ket{i_\up}\bra{i_\up}\otimes \ket{i_\dw}\bra{i_\dw}
  \end{aligned}
\end{equation}
where the constant $C_{i,b}$ take care of the sign change due to
exchanging bath and impurity bit sets. As effect of the Kronecker
delta symbols for bath and impurity these constants are identically 1.  
The numerical implementation relies on the use of the {\tt
  sparse\_map} ($\%sp$) for the groundstate sector as reported in the
listing below:  
\begin{lstlisting}[style=fstyle,numbers=none,basicstyle={\scriptsize\ttfamily}]
do IimpUp=0,2**Norb-1
  do JimpUp=0,2**Norb-1
    !Finding the unique bath states connecting IimpUp and JimpUp
    call sp_return_intersection(sectorI%H(1)%sp,IimpUp,JimpUp,BATHup,lenBATHup)
    if(lenBATHup==0)cycle
    do IimpDw=0,2**Norb-1
      do JimpDw=0,2**Norb-1
        !Finding the unique bath states connecting IimpDw and JimpDw -> BATHdw(:)
        call sp_return_intersection(sectorI%H(2)%sp,IimpDw,JimpDw,BATHdw,lenBATHdw)
        if(lenBATHdw==0)cycle
        do ibUP=1,lenBATHup
          IbathUp = BATHup(ibUP)
          do ibDW=1,lenBATHdw
            IbathDw = BATHdw(ibDW)
            !Allowed spin Fock space Istates:
            !Iup = IimpUp +  2^Norb * IbathUp
            !Idw = IimpDw +  2^Norb * IbathDw
            iUP= binary_search(sectorI%H(1)%map,IimpUp + 2**Norb*IbathUp)
            iDW= binary_search(sectorI%H(2)%map,IimpDw + 2**Norb*IbathDw)
            i  = iUP + (iDW-1)*sectorI%DimUp
            !Allowed spin Fock space Jstates:
            !Jup = JimpUp +  2^Norb * IbathUp
            !Jdw = JimpDw +  2^Norb * IbathDw
            jUP= binary_search(sectorI%H(1)%map,JimpUp + 2**Norb*IbathUp)
            jDW= binary_search(sectorI%H(2)%map,JimpDw + 2**Norb*IbathDw)
            j  = jUP + (jDW-1)*sectorI%DimUp
            ! 
            io = (IimpUp + 2**Norb*IimpDw) + 1
            jo = (JimpUp + 2**Norb*JimpDw) + 1
            irdm(io,jo) = irdm (io,jo) + psi(i)*psi(j)*weight
          enddo
        enddo
      enddo
    enddo
  enddo
enddo
\end{lstlisting}

@@

In presence of these symmetries suitable modifications related to the
absence of Fock space factorization have to be taken into account. To
start with a generic Fock state appearing in the groundstate
decomposition has the form:
$$
\ket{\vec{n}} =\ket{\vec{i}_\up\vec{b}_\up\vec{i}_\dw\vec{b}_\dw}
%
$$
% where the second is just a formal identity.  
Inserting this latter in the definition of iRDM we obtain:
\begin{equation}
  \begin{aligned}
    \Tr_{b_\up, b_\dw}{\rho} &=
    \sum_{b_\sigma=1}^{D_{b_\sigma}}
    \bra{b_\up\, b_\dw}
    \rho
    \ket{b_\dw\, b_\up}    \cr
    &=
    \sum_{b_\sigma=1}^{D_{b_\sigma}}
    \sum_{i_\sigma=1}^{D_{i_\sigma}}
    \sum_{p_\sigma=1}^{D_{p_\sigma}}
    \sum_{j_\sigma=1}^{D_{j_\sigma}} 
    \sum_{q_\sigma=1}^{D_{q_\sigma}}
    a_{i_\up p_\up i_\dw p_\dw} a^*_{j_\up q_\up j_\dw q_\dw}% \cr
    % &{\phantom =}
    \langle{b_\up\, b_\dw}|{i_\up\, p_\up\, i_\dw\, p_\dw}\rangle
    \langle{q_\dw \,  j_\dw \, q_\up\, j_\up }|{b_\dw\, b_\up}\rangle\, .
  \end{aligned}
\end{equation}
In order to simplify this expression we should get rid of the sum with
respect to the bath indices $p_\sigma$ and $q_\sigma$, ideally
contracting them with the outer indices $b_\sigma$.
However, this requires to bring all the bath bits so to preced the
impurity ones:
$$
\ket{i_\up\, p_\up\, i_\dw\, p_\dw} \rightarrow C_{p_\up,i_\dw} \ket{i_\up\, i_\dw \,p_\up\,p_\dw}
$$
where $C_{p_\up,i_\dw}  = (-1)^{\#\vec{n}_{p\up} \cdot
  \#\vec{n}_{i\dw}}$ with   $\#\vec{n}=\sum_i n_i$ is an overall
fermionic sign which depends on the occupation of the $p_\up$ and the
$i_\dw$ bit configurations. Inserting this relation in the equation
for the iRDM we obtain:
\begin{equation}
  \begin{aligned}
    \Tr_{b_\up, b_\dw}{\rho} &=
    \sum_{i_\sigma=1}^{D_{i_\sigma}}
    \sum_{j_\sigma=1}^{D_{j_\sigma}}
    \left[\sum_{b_\sigma=1}^{D_{b_\sigma}}
    a_{i_\up b_\up i_\dw b_\dw} a^*_{j_\up b_\up j_\dw b_\dw}
    C_{b_\up,i_\dw}C_{b_\up,j_\dw}\right]
    \ket{i_\up\, i_\dw}\bra{j_\dw \, j_\up}\cr
&=
    \sum_{i_\sigma=1}^{D_{i_\sigma}}
    \sum_{j_\sigma=1}^{D_{j_\sigma}}
    \rho^{i}_{i_\up i_\dw j_\dw j_\up}
    \ket{i_\up\, i_\dw}\bra{j_\dw \, j_\up} 
  \end{aligned}
\end{equation}
The numerical implementation requires few modifications with respect
to the previous case. In particular it is important to recall that the
ordering of the key-value pairs in the {\tt sparse\_map} in this case
follow a contiguous ordering and that their decomposition is necessary
to reconstruct the correct Fock state index, see the following
listing:
\begin{lstlisting}[style=fstyle,numbers=none,basicstyle={\scriptsize\ttfamily}]
do IimpUp=0,2**Norb-1
  do IimpDw=0,2**Norb-1
    do JimpUp=0,2**Norb-1
      do JimpDw=0,2**Norb-1
        !Build indices of the RDM in 1:4**Norb
        iImp = iImpUp + iImpDw*2**Norb
        jImp = jImpUp + jImpDw*2**Norb
        call sp_return_intersection(sectorI%H(1)%sp,iImp,jImp,Bath,lenBath)
        if(lenBATH==0)cycle
        do ib=1,lenBath
          iBath = Bath(ib)
          !Decompose iBath into B_Up, B_dw to reconstruct B bit set. 
          !B --> {B_up,B_dw}
          iBathUp = mod(iBath,2**Nbath)
          iBathDw = (iBath)/2**Nbath
          !Reconstruct the Fock state Ii map back to sector state i
          ii= iImpUp + iImpDw*2**Ns + 2**Norb*(IBathUp + IBathDw*2**Ns)
          i = binary_search(sectorI%H(1)%map,ii)
          !Reconstruct the Fock state Jj map back to sector state j
          jj= jImpUp + jImpDw*2**Ns + 2**Norb*(IBathUp + IBathDw*2**Ns)
          j = binary_search(sectorI%H(1)%map,jj)
          !Build the signs of each components of RDM(io,jo)
          nBup  = popcnt(Ibits(ii,Norb,Norb*Nbath))
          nIdw  = popcnt(Ibits(ii,Ns,Norb))
          nJdw  = popcnt(Ibits(jj,Ns,Norb))
          signI = (-1)**(nIdw*nBup)
          signJ = (-1)**(nJdw*nBup)
          sgn   = signI*signJ
          !  
          io = (iImpUp+1) + 2**Norb*iImpDw
          jo = (jImpUp+1) + 2**Norb*jImpDw
          rdm(io,jo) = rdm(io,jo) + psi(i)*conjg(psi(j))*weight*sgn
        enddo
      enddo
    enddo
  enddo
enddo
\end{lstlisting}

\subsection{Bath Functions}\label{sSecFunc}
The module {\tt ED\_BATH\_FUNCTIONS} implements on-the-fly
calculations of the hybridization function $\Delta_{\a\b\s\s'}(z)=\sum_\nu
{V^\nu_{\a\s}}(z\11-h^\nu_{\a\b\s\s'})^{-1}V^\nu_{\b\s'}$ and the
non-interacting Green's functions $G^0_{\a\b\s\s'}(z) =
[(z + \mu)\delta_{\a\b\s\s'} - h^0_{\a\b\s\s'} -
\Delta_{\a\b\s\s'}(z)]^{-1}$ for any frequency $z\in\CCC$. 
The module contains implementations for any case according to the
values of {\tt ed\_mode} and {\tt bath\_type}, either with supplied
user bath (a rank-1 array of doubles) or using the allocated {\tt
  effective\_bath} instance {\tt dmft\_bath}. 
We provide different functions to get either $\Delta$, $G^0$ and its
directly its inverse $[G^0]^{-1}$, which are used to evaluate the
self-energy functions upon request, see \secu{sSecIO}.
In the superconductive case {\tt ed\_mode}={\bf superc} different
procedures can be used to retrieve the anomalous components of any
function defined above in the Nambu basis.

\subsection{Bath Optimization}\label{sSecFit}
{\color{red} to be revised, copy\&paste from edipack paper: well it is
  essentiall the same shit} 
In the DMFT context the bath needs to be optimized
against a given realization of the Weiss field
${\GG^{-1}_{0}}_{\a\b\s\s'}(z)$, or the corresponding {\it hybridization} function
$\Theta = (z + \mu)\11 - H^{loc} -\GG_0^{-1}](z)$. While the Weiss
field is obtained from the DMFT self-consistency
equation~\cite{Georges1996RMP}, bath discretization impose to  find an optimal
description of the continuous Weiss field in terms of the finite
number of parameters of the bath (see \secu{sSecBath}.
Although different algorithms can be envisaged to perform this
step~\cite{Garcia2004PRL,Taranto2012PRB} which are perfectly valid in
some physical contexts.
In order to enable the users to use their preferred 
method in \NAME we keep the optimization implementation independent
from the rest of the code. Yet, we offer a fully fledged optimization
strategy based on the  conjugate gradient (CG) minimization of the cost function:
$$
\chi = \sum_{n=1}^{L_{fit}}\frac{1}{W_n}||X(i\omega_n) - X^{QIM}(i\omega_n;\{V,h\})||_q
$$
where the symbol $|| \cdot ||_q$ is a suitably defined $q$-norm in the
matrix function space, $X_{\a\b}={\GG_{0}}_{\a\b\s\s'},\,\Theta_{\a\b\s\s'}$ are the user
supplied local functions and
$X_{\a\b}^{QIM}=G^0_{\a\b\s\s'},\,\Delta_{\a\b\s\s'}$ the
quantum impurity model functions, defined above.
Notably, more sophisticated optimization algorithms have been
recently implemented to better undertake the minimization of similar
cost functions. However, these developments are designed to work with truncated
algorithms, e.g. sCI or DMRG, describing systems with a larger number of
bath levels.   
In the ED context however, where the efficient and wise selection of
the relatively few bath levels becomes crucial, the CG minimization of
the local functions proved to be the most efficient and
flexible optimization scheme so far.

The fit procedure is entirely wrapped by the single function {\tt
  ed\_chi2\_fitgf} contained in the module \texttt{ED\_BATH\_FIT}.   
In order to exploit regularity of the functions the fit is 
performed using imaginary Matsubara frequency. 
The actual form of $X_{\a\b\s\s'}$ is controlled by the input parameter
\texttt{cg\_Scheme=Weiss,Delta}.

In order to fine tune the optimization step in \NAME\ we
introduced a number of control parameters. To start with, we include
two distinct CG algorithms in the library, controlled by the input variable
\texttt{cg\_method=0,1}. The value \texttt{0} corresponds to a
Fletcher-Reeves-Polak-Ribiere minimisation algorithm, adapted from
Numerical Recipes~\cite{NumRec77}. Yet, in order to guarantee
back-compatibility with respect to the past literature, we also provide the
possibility of using the minimization procedure published in
Ref.\onlinecite{Georges1996RMP} which has been largely used in the DMFT
community.

@@

{\bf REMARK TODO}
Notes on the norm function: a) the definition for conventional
matrices, ;b) the definition for the replica/general case where you
can choose between different norm defition.

\subsection{Input/Output}\label{sSecIO}
The module {\tt ED\_IO} provides access to the results of the Lanczos
diagonalization of the quantum impurity problem.
As any single instance of the code is preserved in the memory until a
new call is made, access to the memory pool is only possible through
specific functions implemented in {\tt ED\_IO}. For instance, dynamical response functions, self-energy components or impurity
observables can all be retrieved using dedicated functionalities.
Additionally we provide procedures to perform on-the-fly
re-calculation of the impurity Green's functions and self-energy on a
given arbitrary set of points in the complex frequency domain
exploiting their storage in terms of {\tt gfmatrix} objects.
A long list of available functions is contained in the on-line
documentation (see
\href{https://edipack.github.io/EDIpack2.0/edipack2/13_io.html#ed-io}{https://edipack.github.io/EDIpack2.0/edipack2/13\_io.html\#ed$-$io})
Here we limit the discussion to two paradigmatic examples.
The first is the function {\tt ed\_get\_dens} which write on the
provided input array the value of the orbital impurity occupations
$\langle n_{\a\s}\rangle$. Analogously {\tt ed\_get\_sigma} is used to
retrieve the normal or anomaous components of the Matsubara or
Real-axis self-energy function $\Sigma(z)$. This latter represents the
key output of the calculation in 
any DMFT application.
